Halving the cost of quantum addition

### 2017-09-02 paper.tex
Tweak polish

We also show how to construct an $n$-bit controlled adder circuit with T-count of $8n + O(1)$, and a temporary adder that can be computed for the same cost as the normal adder but whose result can be kept until it is later uncomputed without using T gates.
\end{abstract}

@@

Existing Toffoli constructions use at least 4 T gates per Toffoli gate.
The Cuccaro adder uses $2n-1$ Toffoli gates \citep{Cuccaro2004}.
Correspondingly, existing adder constructions have T-counts of $8n + O(1)$ \citep{Amy2013} and this has been the case for over a decade \citep{Barenco1995, Cuccaro2004}.
The leading factor of 8 has been conjectured to be optimal \citep{AustinDiscussionsAndEmails2017}.

@@

We halve the number of T gates needed to perform an adder by halving the number of T gates needed to perform Toffoli gates that appear in compute/uncompute pairs.
Although we focus on adders in this paper, paired Toffoli gates appear in a wide variety of circuits.
We discuss some specific examples in \autoref{sec:other-applications} (e.g. applying the same rotation $R_Z(\theta)$ to several qubits), but leave further low hanging fruit as an exercise for the reader.

Our paper is divided into four sections.
In \autoref{sec:introduction}, we motivate the need to optimize T-counts, discuss existing work, and note the wide applicability of our low T-count constructions.
\autoref{sec:construction} presents our $4n + O(1)$ T-count adder and explains its pieces, including the temporary logical-AND.
\autoref{sec:other-applications} discusses several circuit constructions and estimates improved by our adder and temporary logical-AND constructions.
Finally, \autoref{sec:conclusion} summarizes our contributions and discusses future work.

@@

Matched Toffolis also appear in quantum circuits rooted in physics instead of abstract mathematics.
For example, we used temporary logical-ANDs to cut the projected T-count of a step in a simulation algorithm under development by nearly 30\% \citep{RyanEmails2017}.
We re-iterate that this is a very widely applicable optimization.

@@

For over a decade, the T-count of addition has been $8n + O(1)$ \citep{Amy2013, Barenco1995, Cuccaro2004}.
In this paper showed how to halve the leading factor of this cost.
Furthermore, we did so via a construction that applies to many other circuits: the temporary logical-AND gate.
Because the temporary logical-AND is so widely applicable, many existing estimates of the T-count of circuits need to be redone.

It is interesting to consider if addition or temporary logical-AND gates can be done with still fewer T gates.
\autoref{fig:lower-bound-logical-AND} presents a simple proof that the T-count of the temporary logical-AND is at least 2.
We expect that the true lower bound is 4, but that there will be special cases with a lower cost.
For example, when a register is conditionally incremented several times, it could be possible to cancel some of the T-count-using computation work at the start of an increment against the uncomputation work at the end of the previous increment.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\qT &\qw && &&&\ctrl{2}&\qw     &\ctrl{3}&\qw     &\qw &\qw     &\ctrl{3}&\qw     &\ctrl{2}&\qw &\\
      &\qT &\qw &&=&&&\targ   &\ctrl{1}&\qw     &\ctrl{1}&\qw &\ctrl{1}&\qw     &\ctrl{1}&\targ   &\qw &\\
      &\qT &\qw && &&&\targ   &\ctrl{1}&\targ   &\targ   &\qT &\targ   &\targ   &\ctrl{1}&\targ   &\qw &\\
      &    &    && &&&        &        &\targ   &\qw     &\qS &\qw     &\targ   &\qw     &        &    &\\
    }
  }
  \caption{
	Three T gates can be synthesized by performing a temporary logical-AND, one T gate, and various Clifford operations.
	Clifford operations can't produce T gates, so the extra T gates must be coming from the temporary logical-AND.
	This proves that at least two T gates went into performing the temporary logical-AND.
  }
  \label{fig:lower-bound-logical-AND}
\end{figure}

Regardless of whether there will be further T-count improvements for an operation as basic as the temporary logical-AND, it is clear that many other T-count optimizations are waiting to be found.
Not just in the construction of basic low-level operations, but in medium-level constructions, high-level constructions, lower-than-circuit level constructions and generally across the whole technology stack that will be needed to perform error corrected quantum computation with the surface code for the first time.

>>>>

We also present an $n$-bit controlled adder circuit with T-count of $8n + O(1)$, a temporary adder that can be computed for the same cost as the normal adder but whose result can be kept until it is later uncomputed without using T gates, and discuss some other constructions whose T-count is improved by the temporary logical-AND.
\end{abstract}

@@

Existing Toffoli constructions use at least 4 T gates per Toffoli gate.
The Cuccaro adder uses $2n + O(1)$ Toffoli gates \citep{Cuccaro2004}.
Correspondingly, existing adder constructions have T-counts of $8n + O(1)$ \citep{Amy2013} and this has been the case for over a decade \citep{Barenco1995, Cuccaro2004}.
The leading factor of 8 has been conjectured to be optimal \citep{AustinDiscussionsAndEmails2017}.

@@

We halve the number of T gates needed to perform an adder by halving the number of T gates needed to perform Toffoli gates that appear in compute/uncompute pairs.
Although we focus on adders in this paper, paired Toffoli gates appear in a wide variety of circuits.
We discuss some specific examples in \autoref{sec:other-applications} (e.g. applying the same rotation $R_Z(\theta)$ to several qubits).

Our paper is divided into four sections.
We motivate the need to optimize T-counts and discuss existing work in \autoref{sec:introduction}.
\autoref{sec:construction} presents our $4n + O(1)$ T-count adder and explains its pieces, including the temporary logical-AND.
In \autoref{sec:other-applications}, we discuss several circuit constructions and estimates improved by our adder and temporary logical-AND constructions.
Finally, \autoref{sec:conclusion} summarizes our contributions and discusses future work.

@@

Matched Toffolis also appear in quantum circuits rooted in physics instead of abstract mathematics.
For example, we used temporary logical-ANDs to cut the projected T-count of a step in a simulation algorithm under development by nearly 30\% \citep{RyanEmails2017}.

@@

For over a decade, the T-count of addition has been $8n + O(1)$ \citep{Amy2013, Barenco1995, Cuccaro2004}.
In this paper we showed how to halve the leading factor of this cost.
Furthermore, we did so via a construction that applies to many other circuits: the temporary logical-AND gate.

Because the temporary logical-AND is so widely applicable, the T-count of many circuits is lower than previously thought.
We covered a few examples in this paper but, for brevity, we leave further low hanging fruit as an exercise for the reader.

It is interesting to consider if addition or temporary logical-AND gates can be done with still fewer T gates.
\autoref{fig:lower-bound-logical-AND} presents a simple proof that the T-count of the temporary logical-AND is at least 2.
Regardless of whether the true lower bound is 4, we expect that there are special cases with lower average T-counts.
For example, consider a register being conditionally incremented several times.
Incrementing is performed with paired Toffoli gates that can be replaced by temporary logical-ANDs.
An alternative way to reduce the T-count is by cancelling computation work at the start of an increment against uncomputation work from the end of the previous increment.
The alternative approach happens to have the same T-count in this case, but that may not always be the case.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\qT &\qw && &&&\ctrl{2}&\qw     &\ctrl{3}&\qw     &\qw &\qw     &\ctrl{3}&\qw     &\ctrl{2}&\qw &\\
      &\qT &\qw &&=&&&\targ   &\ctrl{1}&\qw     &\ctrl{1}&\qw &\ctrl{1}&\qw     &\ctrl{1}&\targ   &\qw &\\
      &\qT &\qw && &&&\targ   &\ctrl{1}&\targ   &\targ   &\qT &\targ   &\targ   &\ctrl{1}&\targ   &\qw &\\
      &    &    && &&&        &        &\targ   &\qw     &\qS &\qw     &\targ   &\qw     &        &    &\\
    }
  }
  \caption{
	Three T gates can be synthesized by performing a temporary logical-AND, one T gate, and various Clifford operations.
	Clifford operations can't produce T gates, so the extra T gates must be coming from the temporary logical-AND.
	This proves that performing the temporary logical-AND requires at least two T gates.
  }
  \label{fig:lower-bound-logical-AND}
\end{figure}

Regardless of whether there will be further T-count improvements for operations as basic as the temporary logical-AND, it is clear that many other T-count optimizations are waiting to be found.
Not just in the construction of basic low-level operations, but in medium-level constructions, high-level constructions, lower-than-circuit level constructions, and generally across the whole technology stack that will be needed to perform error corrected quantum computation with the surface code for the first time.
